Tutkielman otsikko

### 2020-02-06 gradu/msc_english_contents.tex
more stuff

However, random access of compressed data is needed in compressed data structures. In most compression methods, the only way 
to this is to decompress data from the beginning. An integer compressing method with fast random access is explained and compared
existing state-of-the-art methods.

@@

Variable-byte encoding \citep{Wil99} (VB) is a method of compressing integers via omitting leading zero bits. A good data set for 
VB encoding is a list of different sized numbers. As an example, an inverted index used by document search engine stores word frequencies and 
locations in documents. VB saves space here because index numbers need to support large values, but are usually significantly smaller. 
It is also proven to improve search speed, because less bytes need to be read from a hard drive \citep{Sch02}.

VB splits an integer into blocks of $b$ bits. A continuation bit is added to the block to form a chunk $c$ and then chunk non-empty chunks are stored.
The continuation bit is set to 1 for the last block of the integer. During decoding, chunks are read and their blocks are joined until a continuation 
bit is set to 1. An example implementation is showcased in Figure~\ref{fig:vbyte}.

\begin{figure}[h!] 
\centering 
blblblblb
testataan
tätä
\caption{VB encoding and decoding.\label{fig:vbyte}}
\end{figure}

Small lengths of $c$ can yield better
compression rate at the cost of more bit manipulation, while longer chunks need less bit manipulation and offer less effective compression. 
Generally chunk length of 8 has been used because it gives a good average and handling bytes is convenient \citep{IRbook}.

bl 
It splits an integer into blocks of $b$ bits and then encodes
the block as $b + c$ where $c$ is a bit denoting whether $b$ has the most significant bits of the number or not.

This method loses in compression performance to other methods \citep{Bri09}, but decoding is fast.

@@

\chapter{Results}
 - comparison to basic implementation + Bri09
\begin{table}
\centering
\caption{Results with 100k entries (in milliseconds).\label{table:results}}
\begin{tabular}{l||l c c c c r} 
Experiment & 128 & 256 & 32768 & 65536 & $2^{24}$ & $2^{32} -1$\\ 
\hline \hline 
$7bit VByte encoding$ & 34.97 & 49 & 53.04 & 52.18 & 53.08 & 76.21\\
$8bit VByte encoding$ & 33.57 & 32.47 & 42.96 & 43.11 & 46.15 & 65.14\\
$7bit VByte encoding with array$ & 33.39 & 46.85 & 51.24 & 49.03 & 48.93 & 66.84 \\
$8bit VByte encoding with array$ & 32.52 & 31.88 & 41.54 & 39.94 & 41.15 & 52.86 \\

@@

\begin{table}
\centering
\caption{Results with 1M entries (in milliseconds).\label{table:results}}
\begin{tabular}{l||l c c c c r} 
Experiment & 128 & 256 & 32768 & 65536 & $2^{24}$ & $2^{32} -1$ \\ 
\hline \hline 
$7bit VByte encoding$ & 38.17 & 55.09 & 64.38 & 65.36 & 68.08 & 159 \\
$8bit VByte encoding$ & 37.09 & 37.75 & 53.44 & 54.6 & 59.32 & 148.7\\
$7bit VByte encoding with array$ & 38.09 & 55.42 & 62.22 & 61.25 & 71.72 & 135.01\\
$8bit VByte encoding with array$ & 36.13 & 36.83 & 50.58 & 50.73 & 56.93 & 103.18\\

@@

\section{Figures}
Figure~\ref{fig:logo} gives an example how to add figures to the document. Remember always to cite the figure in the main text.

\begin{figure}[h!] 
\centering 
\includegraphics[width=0.3\textwidth]{HY-logo-ml.png}
\caption{University of Helsinki flame-logo for Faculty of Science.\label{fig:logo}}
\end{figure}

@@

Table~\ref{table:results} gives an example how to report experimental results. Remember always to cite the table in the main text.

\chapter{Citations}

\section{Citations to literature}

References are listed in a separate .bib-file. In this case it is named \texttt{bibliography.bib} including the following content:
\begin{verbatim}

@article{einstein,
    author =       "Albert Einstein",
    title =        "{Zur Elektrodynamik bewegter K{\"o}rper}. ({German})
        [{On} the electrodynamics of moving bodies]",
    journal =      "Annalen der Physik",
    volume =       "322",
    number =       "10",
    pages =        "891--921",
    year =         "1905",
    DOI =          "http://dx.doi.org/10.1002/andp.19053221004"
}

@book{latexcompanion,
    author    = "Michel Goossens and Frank Mittelbach and Alexander Samarin",
    title     = "The \LaTeX\ Companion",
    year      = "1993",
    publisher = "Addison-Wesley",
    address   = "Reading, Massachusetts"
}

@misc{knuthwebsite,
    author    = "Donald Knuth",
    title     = "Knuth: Computers and Typesetting",
    url       = "http://www-cs-faculty.stanford.edu/%7Eknuth/abcde.html"
}
\end{verbatim}

In the last reference url field the code \verb+%7E+ will translate into \verb+~+ once clicked in the final pdf.

References are created using command \texttt{\textbackslash cite\{einstein\}}, showing as \citep{einstein}. Other examples: \citep{latexcompanion,knuthwebsite}.

Citation style can be negotiated with the supervisor. See some options in \url{https://www.sharelatex.com/learn/Bibtex_bibliography_styles}.

\section{Crossreferences}

Appendix~\ref{appendix:model} on page~\pageref{appendix:model} contains some additional material.

>>>>

However, random access of compressed data is needed in compressed data structures. In most compression methods, the only way 
to this is to decompress data from the beginning. An integer compressing method with fast random access is explained later and compared
existing state-of-the-art methods.

@@

Variable-byte encoding \citep{Wil99} (VB) is a method of compressing integers via omitting leading zero bits. In normal data sets, it loses in 
compression performance to generic algorithms like Huffman encoding or Lempel-Ziv encoding, but in comparison it's faster to decode \citep{Bri09} 
and allows constant time random access (TODO: link to later chapters). A good data set for VB encoding is a list of mostly small numbers with a 
need to support larger ones.

A search engine may use an inverted index of words in documents. For each word, a list of document IDs where the word appears is stored. 
It may also store locations of the word in document for advanced search purposes. Usually these lists are preprocessed and stored 
as gaps, storing the difference to previous number instead of the actual number \citep{Man08}. Common words have a lot of entries in these lists, 
but because of gap storing the numbers are small. In contrast, rare words have only a few entries but the numbers stored are larger. These lists 
are excellent data sets for VB. Apache Lucene has variable bytes as vInt and IBM DB2 has them as Variable Byte).

TODO: a chapter about existing work
Usage of VB to compress inverted 
index lists was first experimented in \citep{Sch02}. (TODO: find better examples or find references to these)

VB splits each integer into blocks of $b$ bits and adds a continuation bit to the block to form chunks of length $1+b$. The extra bit is set to 1 only
on the block containing the least significant bits of the integer. This information is used in decoding to signal if next chunk continues the current 
integer. For example, let's assume $b = 4$ and let $n = 42$ be a 16 bit unsigned integer. The standard 8-bit representation of $n$ is 
\texttt{00000000 00101000}. When split to blocks of $b$ bits, it becomes \texttt{0000 0000 0010 1000}. Empty blocks are omitted and continuation bits 
are added to the remaining blocks. The result is \texttt{00010 11000}, which is the compressed data.

\medskip
\noindent\begin{minipage}{.5\textwidth}
\begin{algorithmic}
\Function{VBEncodeNumber}{$n$}
\State $bytes\gets $ list
\While{true}
\State $bytes$.prepend($n \bmod 128$)
\If{$n < 128$} \State break \EndIf
\State $n\gets n$ div $128$
\EndWhile
\State $bytes$.last() += $128$
\State \textbf{return} bytes
\EndFunction
\captionof{figure}{VByte encoding} \label{vbyte_enc}
\end{algorithmic}
\end{minipage}%
\begin{minipage}{.5\textwidth}
\begin{algorithmic}
\Function{VBEncode}{$numbers$}
\State $bytestream\gets $ list
\ForEach {$n \in numbers$}
\State $bytes \gets$ VBEncodeNumber($n$)
\State $bytestream$.extend($bytes$)
\EndFor
\State \textbf{return} bytestream
\EndFunction
\end{algorithmic}
\end{minipage}
\medskip

Decoding is essentially just reversing the encoding steps: chunks are read until a chunk with 1 as continuation bit is found. Continuation bits 
are removed from all the chunks and the blocks are concatenated to form the original number. As in the previous example, $b = 4$, encoded message 
is \texttt{00010 11000} and $n = 0$. Block from the first chunk is extracted and added to $n$, making $n = $\texttt{10}. A bitwise shift to left 
equal to $b$ is applied to $n$, changing $n = $\texttt{100000}. Then the block is extracted from the next chunk. This block is added  $n$, making 
$n = 42$. Because the previous continuous bit was 1, decoding this number has finished. An example implementation of encode and decode with block 
length of 7 is shown in Figure~\ref{vbyte_enc} and Figure~\ref{vbyte_dec}.

Small lengths of $c$ can yield better compression rate at the cost of more bit manipulation, while longer chunks need less bit manipulation and 
offer less effective compression. Generally block length of 7 has been used because it gives a good average and handling chunks as bytes is 
convenient \citep{Man08}.

\begin{figure}[ht]
\begin{algorithmic}
\Function{VBDecode}{$bytestream$}
\State $numbers\gets $ list
\State $n\gets 0$
\ForEach {$b \in bytestream$}
\If{$b < 128$}
\State $n\gets 128\times n $ + $b$
\Else
\State $n\gets 128\times n $ + $b$ - $128$
\State $numbers$.append($n$)
\State $n\gets 0$
\EndIf
\EndFor
\State \textbf{return} numbers
\EndFunction
\end{algorithmic}
\caption{VByte decoding} \label{vbyte_dec}
\end{figure}

+description

\medskip
-motivation (see papers below)

\medskip
+pseudocode

\medskip
+a few examples (1)

\medskip
-a review of related work should include:

\medskip
---the text book from which the initial description I sent you comes from

\medskip
---First use for inverted index compression (at least that I'm aware of):

\medskip
https://dl.acm.org/doi/10.1145/564376.564416

\medskip
---More recent incarnations of vbyte (just mention them):

\medskip
https://arxiv.org/abs/1709.08990

\medskip
https://arxiv.org/abs/1503.07387

\medskip

@@

\chapter{Results}
 - comparison to basic implementation + Bri09
\begin{table}
\centering
\caption{Results with 100k entries (in milliseconds).\label{table:results1}}
\begin{tabular}{l||l c c c c r} 
Experiment & 128 & 256 & 32768 & 65536 & $2^{24}$ & $2^{32} -1$\\ 
\hline \hline 
$7bit VByte encoding$ & 34.97 & 49 & 53.04 & 52.18 & 53.08 & 76.21\\
$8bit VByte encoding$ & 33.57 & 32.47 & 42.96 & 43.11 & 46.15 & 65.14\\
$7bit VByte encoding with array$ & 33.39 & 46.85 & 51.24 & 49.03 & 48.93 & 66.84 \\
$8bit VByte encoding with array$ & 32.52 & 31.88 & 41.54 & 39.94 & 41.15 & 52.86 \\

@@

\begin{table}
\centering
\caption{Results with 1M entries (in milliseconds).\label{table:results2}}
\begin{tabular}{l||l c c c c r} 
Experiment & 128 & 256 & 32768 & 65536 & $2^{24}$ & $2^{32} -1$ \\ 
\hline \hline 
$7bit VByte encoding$ & 38.17 & 55.09 & 64.38 & 65.36 & 68.08 & 159 \\
$8bit VByte encoding$ & 37.09 & 37.75 & 53.44 & 54.6 & 59.32 & 148.7\\
$7bit VByte encoding with array$ & 38.09 & 55.42 & 62.22 & 61.25 & 71.72 & 135.01\\
$8bit VByte encoding with array$ & 36.13 & 36.83 & 50.58 & 50.73 & 56.93 & 103.18\\

@@

\section{Figures}
Figure gives an example how to add figures to the document. Remember always to cite the figure in the main text.

@@

Table gives an example how to report experimental results. Remember always to cite the table in the main text.

### 2020-02-06 gradu/temp/msc_english_contents.tex
more stuff

However, random access of compressed data is needed in compressed data structures. In most compression methods, the only way 
to this is to decompress data from the beginning. An integer compressing method with fast random access is explained later and compared
existing state-of-the-art methods.

@@

Variable-byte encoding \citep{Wil99} (VB) is a method of compressing integers via omitting leading zero bits. In normal data sets, it loses in 
compression performance to generic algorithms like Huffman encoding or Lempel-Ziv encoding, but in comparison it's faster to decode \citep{Bri09} 
and allows constant time random access (TODO: link to later chapters). A good data set for VB encoding is a list of mostly small numbers with a 
need to support larger ones.

A search engine may use an inverted index of words in documents. For each word, a list of document IDs where the word appears is stored. 
It may also store locations of the word in document for advanced search purposes. Usually these lists are preprocessed and stored 
as gaps, storing the difference to previous number instead of the actual number \citep{Man08}. Common words have a lot of entries in these lists, 
but because of gap storing the numbers are small. In contrast, rare words have only a few entries but the numbers stored are larger. These lists 
are excellent data sets for VB. Apache Lucene has variable bytes as vInt and IBM DB2 has them as Variable Byte).

TODO: a chapter about existing work
Usage of VB to compress inverted 
index lists was first experimented in \citep{Sch02}. (TODO: find better examples or find references to these)

VB splits each integer into blocks of $b$ bits and adds a continuation bit to the block to form chunks of length $1+b$. The extra bit is set to 1 only
on the block containing the least significant bits of the integer. This information is used in decoding to signal if next chunk continues the current 
integer. For example, let's assume $b = 4$ and let $n = 42$ be a 16 bit unsigned integer. The standard 8-bit representation of $n$ is 
\texttt{00000000 00101000}. When split to blocks of $b$ bits, it becomes \texttt{0000 0000 0010 1000}. Empty blocks are omitted and continuation bits 
are added to the remaining blocks. The result is \texttt{00010 11000}, which is the compressed data.

\medskip
\noindent\begin{minipage}{.5\textwidth}
\begin{algorithmic}
\Function{VBEncodeNumber}{$n$}
\State $bytes\gets $ list
\While{true}
\State $bytes$.prepend($n \bmod 128$)
\If{$n < 128$} \State break \EndIf
\State $n\gets n$ div $128$
\EndWhile
\State $bytes$.last() += $128$
\State \textbf{return} bytes
\EndFunction
\captionof{figure}{VByte encoding} \label{vbyte_enc}
\end{algorithmic}
\end{minipage}%
\begin{minipage}{.5\textwidth}
\begin{algorithmic}
\Function{VBEncode}{$numbers$}
\State $bytestream\gets $ list
\ForEach {$n \in numbers$}
\State $bytes \gets$ VBEncodeNumber($n$)
\State $bytestream$.extend($bytes$)
\EndFor
\State \textbf{return} bytestream
\EndFunction
\end{algorithmic}
\end{minipage}
\medskip

Decoding is essentially just reversing the encoding steps: chunks are read until a chunk with 1 as continuation bit is found. Continuation bits 
are removed from all the chunks and the blocks are concatenated to form the original number. As in the previous example, $b = 4$, encoded message 
is \texttt{00010 11000} and $n = 0$. Block from the first chunk is extracted and added to $n$, making $n = $\texttt{10}. A bitwise shift to left 
equal to $b$ is applied to $n$, changing $n = $\texttt{100000}. Then the block is extracted from the next chunk. This block is added  $n$, making 
$n = 42$. Because the previous continuous bit was 1, decoding this number has finished. An example implementation of encode and decode with block 
length of 7 is shown in Figure~\ref{vbyte_enc} and Figure~\ref{vbyte_dec}.

Small lengths of $c$ can yield better compression rate at the cost of more bit manipulation, while longer chunks need less bit manipulation and 
offer less effective compression. Generally block length of 7 has been used because it gives a good average and handling chunks as bytes is 
convenient \citep{Man08}.

\begin{figure}[ht]
\begin{algorithmic}
\Function{VBDecode}{$bytestream$}
\State $numbers\gets $ list
\State $n\gets 0$
\ForEach {$b \in bytestream$}
\If{$b < 128$}
\State $n\gets 128\times n $ + $b$
\Else
\State $n\gets 128\times n $ + $b$ - $128$
\State $numbers$.append($n$)
\State $n\gets 0$
\EndIf
\EndFor
\State \textbf{return} numbers
\EndFunction
\end{algorithmic}
\caption{VByte decoding} \label{vbyte_dec}
\end{figure}

+description

\medskip
-motivation (see papers below)

\medskip
+pseudocode

\medskip
+a few examples (1)

\medskip
-a review of related work should include:

\medskip
---the text book from which the initial description I sent you comes from

\medskip
---First use for inverted index compression (at least that I'm aware of):

\medskip
https://dl.acm.org/doi/10.1145/564376.564416

\medskip
---More recent incarnations of vbyte (just mention them):

\medskip
https://arxiv.org/abs/1709.08990

\medskip
https://arxiv.org/abs/1503.07387

\medskip

@@

\chapter{Results}
 - comparison to basic implementation + Bri09
\begin{table}
\centering
\caption{Results with 100k entries (in milliseconds).\label{table:results1}}
\begin{tabular}{l||l c c c c r} 
Experiment & 128 & 256 & 32768 & 65536 & $2^{24}$ & $2^{32} -1$\\ 
\hline \hline 
$7bit VByte encoding$ & 34.97 & 49 & 53.04 & 52.18 & 53.08 & 76.21\\
$8bit VByte encoding$ & 33.57 & 32.47 & 42.96 & 43.11 & 46.15 & 65.14\\
$7bit VByte encoding with array$ & 33.39 & 46.85 & 51.24 & 49.03 & 48.93 & 66.84 \\
$8bit VByte encoding with array$ & 32.52 & 31.88 & 41.54 & 39.94 & 41.15 & 52.86 \\

@@

\begin{table}
\centering
\caption{Results with 1M entries (in milliseconds).\label{table:results2}}
\begin{tabular}{l||l c c c c r} 
Experiment & 128 & 256 & 32768 & 65536 & $2^{24}$ & $2^{32} -1$ \\ 
\hline \hline 
$7bit VByte encoding$ & 38.17 & 55.09 & 64.38 & 65.36 & 68.08 & 159 \\
$8bit VByte encoding$ & 37.09 & 37.75 & 53.44 & 54.6 & 59.32 & 148.7\\
$7bit VByte encoding with array$ & 38.09 & 55.42 & 62.22 & 61.25 & 71.72 & 135.01\\
$8bit VByte encoding with array$ & 36.13 & 36.83 & 50.58 & 50.73 & 56.93 & 103.18\\

@@

Table gives an example how to report experimental results. Remember always to cite the table in the main text.

>>>>

However, random access of compressed data is needed in compressed data structures. In most compression methods, the only way 
to this is to decompress data from the beginning. An integer compressing method with fast random access is explained and compared
existing state-of-the-art methods.

@@

Variable-byte encoding \citep{Wil99} (VB) is a method of compressing integers via omitting leading zero bits. A good data set for 
VB encoding is a list of different sized numbers. As an example, an inverted index used by document search engine stores word frequencies and 
locations in documents. VB saves space here because index numbers need to support large values, but are usually significantly smaller. 
It is also proven to improve search speed, because less bytes need to be read from a hard drive \citep{Sch02}.

VB splits an integer into blocks of $b$ bits. A continuation bit is added to the block to form a chunk $c$ and then chunk non-empty chunks are stored.
The continuation bit is set to 1 for the last block of the integer. During decoding, chunks are read and their blocks are joined until a continuation 
bit is set to 1. An example implementation is showcased in Figure~\ref{fig:vbyte}.

\begin{algorithm}
\KwData{this text}
\KwResult{how to write algorithm with \LaTeX2e}
\LinesNumbered
\While{not at end of this document}{read current\;
  \eIf{understand}{
  go to next section\;
  current section becomes this one\;
  }{
    go back to the beginning of current section\;
  }
}
\caption{VB encoding.\label{fig:vbyte}}
\end{algorithm}

Small lengths of $c$ can yield better
compression rate at the cost of more bit manipulation, while longer chunks need less bit manipulation and offer less effective compression. 
Generally chunk length of 8 has been used because it gives a good average and handling bytes is convenient \citep{IRbook}.

bl 
It splits an integer into blocks of $b$ bits and then encodes
the block as $b + c$ where $c$ is a bit denoting whether $b$ has the most significant bits of the number or not.

This method loses in compression performance to other methods \citep{Bri09}, but decoding is fast.

@@

\chapter{Results}
 - comparison to basic implementation + Bri09
\begin{table}
\centering
\caption{Results with 100k entries (in milliseconds).\label{table:results}}
\begin{tabular}{l||l c c c c r} 
Experiment & 128 & 256 & 32768 & 65536 & $2^{24}$ & $2^{32} -1$\\ 
\hline \hline 
$7bit VByte encoding$ & 34.97 & 49 & 53.04 & 52.18 & 53.08 & 76.21\\
$8bit VByte encoding$ & 33.57 & 32.47 & 42.96 & 43.11 & 46.15 & 65.14\\
$7bit VByte encoding with array$ & 33.39 & 46.85 & 51.24 & 49.03 & 48.93 & 66.84 \\
$8bit VByte encoding with array$ & 32.52 & 31.88 & 41.54 & 39.94 & 41.15 & 52.86 \\

@@

\begin{table}
\centering
\caption{Results with 1M entries (in milliseconds).\label{table:results}}
\begin{tabular}{l||l c c c c r} 
Experiment & 128 & 256 & 32768 & 65536 & $2^{24}$ & $2^{32} -1$ \\ 
\hline \hline 
$7bit VByte encoding$ & 38.17 & 55.09 & 64.38 & 65.36 & 68.08 & 159 \\
$8bit VByte encoding$ & 37.09 & 37.75 & 53.44 & 54.6 & 59.32 & 148.7\\
$7bit VByte encoding with array$ & 38.09 & 55.42 & 62.22 & 61.25 & 71.72 & 135.01\\
$8bit VByte encoding with array$ & 36.13 & 36.83 & 50.58 & 50.73 & 56.93 & 103.18\\

@@

Table~\ref{table:results} gives an example how to report experimental results. Remember always to cite the table in the main text.

\chapter{Citations}

\section{Citations to literature}

References are listed in a separate .bib-file. In this case it is named \texttt{bibliography.bib} including the following content:
\begin{verbatim}

@article{einstein,
    author =       "Albert Einstein",
    title =        "{Zur Elektrodynamik bewegter K{\"o}rper}. ({German})
        [{On} the electrodynamics of moving bodies]",
    journal =      "Annalen der Physik",
    volume =       "322",
    number =       "10",
    pages =        "891--921",
    year =         "1905",
    DOI =          "http://dx.doi.org/10.1002/andp.19053221004"
}

@book{latexcompanion,
    author    = "Michel Goossens and Frank Mittelbach and Alexander Samarin",
    title     = "The \LaTeX\ Companion",
    year      = "1993",
    publisher = "Addison-Wesley",
    address   = "Reading, Massachusetts"
}

@misc{knuthwebsite,
    author    = "Donald Knuth",
    title     = "Knuth: Computers and Typesetting",
    url       = "http://www-cs-faculty.stanford.edu/%7Eknuth/abcde.html"
}
\end{verbatim}

In the last reference url field the code \verb+%7E+ will translate into \verb+~+ once clicked in the final pdf.

References are created using command \texttt{\textbackslash cite\{einstein\}}, showing as \citep{einstein}. Other examples: \citep{latexcompanion,knuthwebsite}.

Citation style can be negotiated with the supervisor. See some options in \url{https://www.sharelatex.com/learn/Bibtex_bibliography_styles}.

\section{Crossreferences}

Appendix~\ref{appendix:model} on page~\pageref{appendix:model} contains some additional material.
